Quantum H theorem, variational entropy, and relaxation processes

### 2020-12-16 tex/classical.tex
Typos, 1st intro paragraph, secs. 2.0, 2.1

\section{Classical scheme}

@@

In the classical version, the Boltzmann transport equation is required to show
the $H$-theorem. That equation is
%
\begin{equation}\label{eq:transport}
  \left(
    \frac{\partial}{\partial t}+\vec{v}_1 \cdot\nabla_r
    +\frac{\vec{F}}{m}\cdot\nabla_{v_1}
  \right)f_1=\int\mathrm{d}\Omega\int
    \mathrm{d}^{3}v_2\sigma(\Omega)|\vec{v}_1-\vec{v}_2|(f_2'f_1'-f_2f_1),
\end{equation}
%
where $\Omega$ is the solid angle, $\sigma$ is the scattering section, $F$ the
external force applied to the system, $f_1$, $f_2$ are the distribution
functions of the particles 1 and 2 respectively before the collision,
and $f'_1$, $f'_2$ are the distribution function of the particles 1 and 2
respectively after the collision.

On the one hand, dynamics, inhomogeneities, and external forces are described
by the left-hand side of the equation (\ref{eq:transport}). On the other hand, the
binary collisions among particles are involved in the right-hand side of the
previous equation. However, the \textit{molecular chaos hypothesis} was applied
to obtain the right-hand side of (\ref{eq:transport}). The \textit{molecular chaos
hypothesis} establishes that spatial and velocity coordinates of the particles
are not correlated.

%-------------------------
\subsection{H-theorem and the Maxwell-Boltzmann distribution}
The $H$ theorem can be proved using the $H$ functional defined by Boltzmann 
%
\begin{equation}\label{eq:hfunctional}
   H=\int f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v,
\end{equation} 
%
where $f(\vec{v},t)$ is a homogeneous distribution function.
This $H$ functional describes a system with a global energy $E$, a global
number of free classical particles $N$, global volume $V$, and a temperature $T$.
Besides, the Boltzmann's $H$ functional correspond to a density of entropy
when the system arrives at the equilibrium state.

The $H$ theorem establishes that if $f(\vec{v},t)$ satisfies the Boltzmann
transport equation and the molecular chaos hypothesis, then
$dH/dt\leq0$ and the system arrives to the equilibrium state when
$dH/dt=0$.
The theorem can be proved by applying the time derivative
to Eq. (\ref{eq:hfunctional}) and using the Boltzmann transport equation
(\ref{eq:transport}). As a consequence of this theorem, the distribution function
that describes the system in equilibrium corresponds to the Maxwell-Boltzmann
distribution. This distribution can be obtained by solving a sufficient
condition for $f$ to obtain the equilibrium distribution function $f_0$:
%
\begin{equation}
    f_0(\vec{v}'_2)f_0(\vec{v}'_1)- f_0(\vec{v}_2)f_0(\vec{v}_1)=0.
\end{equation}
%

@@

The $H$-theorem fixes the time evolution direction of systems in nature, such
as the second law of thermodynamics does.\footnote{The molecular hypothesis is %
fundamental to determine the physical meaning of the $H$ theorem. For more %
information, see \cite{bib:huang}.}

@@

The Boltzmann $H$-functional (\ref{eq:hfunctional}) is defined with a spatially
homogeneous distribution $f(\vec{v},t)$. However, we want to consider a $H$
functional with a non-homogeneous distribution function in order to generalize
the Boltzmann $H$ theorem. Because of this, we introduce a modified $H$
functional (denoted as $H'$) defined as
%
\begin{equation}\label{eq:CH2}
   H'(t)=\sum_{M=1}^{K}\int f_M(\vec{v},t) \ln f_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%
This $H'$ describes a system with a global energy $E$, a global number of free
classical particles $N$, global volume $V$, and a temperature $T$. Nevertheless,
the global system, in turn, consists of a set of $K$ small cells with a constant
volume $\delta V_M$, whose local homogeneous distribution functions is
$f_{M}(\vec{v},t)$.

@@

To prove the Boltzmann's $H$ theorem (using (\ref{eq:CH2}) as functional), we need
to assume that the distribution function $f_M(\vec{v},t)$ must satisfy the
following assumptions:
%
\begin{itemize}
  \item The local equilibrium hypothesis in each cell ($f_{M}$ must be
    spatial-homogeneous for each $M$).
  \item The non-homogeneous distribution function assumption in the total
    volume ($f_{M}$ must be different among all cells).
  \item The following restrictions
    %
    \begin{equation}\label{restrictionoutsideclassical}
      \int f_{M}(\vec{v},t) \mathrm{d}^{3}v=
      \bar{\mathcal{N}}+\Delta_M(t); \ \ \ \ 
      \int f_{M}(\vec{v},t) \epsilon(\vec{v}) \mathrm{d}^{3}v=
      \bar{\mathcal{E}}+ \delta_M(t),
    \end{equation}
    %
    where $\bar {\mathcal{N}}$ and $\bar{\mathcal{E}}$ are
    the local particle number and the local energy in equilibrium
    %
    \begin{equation}
      \bar{\mathcal{N}}=
      \int \bar{f}(\vec{v}) \mathrm{d}^{3}v ; \quad
      \bar{\mathcal{E}}=
      \int \bar{f}(\vec{v})\epsilon(\vec{v}) \mathrm{d}^{3}v,
    \end{equation}
    %
    and $\Delta_M$, $\delta_M$ could be seen as a deviation from
    $\bar{\mathcal{N}}$ and $\bar{\mathcal{E}}$ respectively,
    with $\Delta_M(t)\ll\bar{\mathcal{N}}$ and
    $\delta_M(t)\ll\bar{\mathcal{E}}$.
\end{itemize} 
%
Those conditions described a system out of equilibrium but no so far from it,
that is a perturbed system and will suffer a \textit{relaxation process, i.e.}
The time evolution of a perturbated system to the equilibrium state.

@@

Joining both results, we can say the following statement:
Consider a classical gas in a total volume $V$ (divided in $K$ cells of equal
volume elements), total energy $E$ and total number of free particles $N$.
Consider also the system has inhomogeneities and suffers a relaxation process.
If we define the following functional
%
\begin{equation}\label{CH3}
   H'(t)=\sum_{M=1}^{K}\int f_M(\vec{v},t) \ln f_M(\vec{v},t)\mathrm{d}^3v,
\end{equation}
%
where $f_M(\vec{v},t)$ is the local distribution function of each cell in the
system, then in the first-order approximation
%
\begin{equation}
    \frac{dH'}{dt} \leq 0.
\end{equation}
%
This statement will be the classical $H$ theorem with inhomogeneities.

In the next section, we present a proposal quantum version of the
$H$ theorem defined by Tolman and our proposal quantum version of the $H$
theorem applying the method of the volume divided into cells.

>>>>

\section{Classical scheme}\label{sec:classical}

Classical statistical mechanics is theoretically founded on the Boltzmann transport
equation and the hypo of molecular chaos. The fundamental connection
between the microscopic nature of matter and the phenomenological laws of thermodynamics
is the $H$-theorem.

@@

From the Boltzmann kinetic theory of gases, the probability for a particle to be in a given
state of position and velocity (\textit{aka} the distribution function) must satify
the transport equation (Eq.~(\ref{eq:transport}), below).
This equation is traditionally obtained by exploiting the volume-invariance of the $\mu$-space
(\textit{i.e.} the space formed by the direct product of position and velocity)
and considering that elastic collisions constitute the predominant mechanism
for exchanging energy among particles \cite{bib:huang}.
%
\begin{equation}\label{eq:transport}
	%\left(\frac{\partial f_1}{\partial t}\right)_{\mathrm{coll}}=
  \left(
    \frac{\partial}{\partial t}+\vec{v}_1 \cdot\nabla_{\vec r}
    +\frac{\vec{F}}{m}\cdot\nabla_{\vec v_1}
  \right)f_1=\int\mathrm{d}\Omega\int
    \mathrm{d}^{3}v_2\sigma(\Omega)|\vec{v}_1-\vec{v}_2|(f_2'f_1'-f_2f_1).
\end{equation}
%
In Eq. (\ref{eq:transport}), $\Omega$ is the solid angle, $\sigma$ is the
scattering cross section, $\vec F$ the
external force applied to the system, and $f_1$ and $f_2$ ($f'_1$ and $f'_2$) are the distribution
functions of particles 1 and 2, respectively, before (after) the collision.

Particle dynamics and external forces are described
by the left-hand side of Eq.~(\ref{eq:transport}), and the right-hand side
is derived considering binary collisions between particles and
the \textit{molecular chaos hypothesis}, \textit{i.e.} it is assumed that
the coordinates and velocities of the particles are not time-correlated.

%-------------------------
\subsection{H-theorem and the Maxwell-Boltzmann distribution}
The $H$-theorem can be proved using the $H$ functional defined by Boltzmann 
%
\begin{equation}\label{eq:hfunctional}
   H=\int f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v,
\end{equation} 
%
where $f(\vec{v},t)$ is a homogeneous distribution function.
This $H$ functional describes a system with a global energy $E$, a global
number of free classical particles $N$, global volume $V$, and a temperature $T$.
Besides, the Boltzmann's $H$ functional correspond to a density of entropy
when the system arrives at the equilibrium state.

The $H$-theorem establishes that if $f(\vec{v},t)$ satisfies the Boltzmann
transport equation and the molecular chaos hypothesis, then
$dH/dt\leq0$ and the system arrives to the equilibrium state when
$dH/dt=0$.
The theorem can be proved by applying the time derivative
to Eq. (\ref{eq:hfunctional}) and using the Boltzmann transport equation
(\ref{eq:transport}). As a consequence of this theorem, the distribution function
that describes the system in equilibrium corresponds to the Maxwell-Boltzmann
distribution. This distribution can be obtained by solving a sufficient
condition for $f$ to obtain the equilibrium distribution function $f_0$:
%
\begin{equation}
    f_0(\vec{v}'_2)f_0(\vec{v}'_1)- f_0(\vec{v}_2)f_0(\vec{v}_1)=0.
\end{equation}
%

@@

The $H$-theorem fixes the time evolution direction of systems in nature, such
as the second law of thermodynamics does.\footnote{The molecular hypothesis is %
fundamental to determine the physical meaning of the $H$-theorem. For more %
information, see \cite{bib:huang}.}

@@

The Boltzmann $H$-functional (\ref{eq:hfunctional}) is defined with a spatially
homogeneous distribution $f(\vec{v},t)$. However, we want to consider a $H$
functional with a non-homogeneous distribution function in order to generalize
the Boltzmann $H$-theorem. Because of this, we introduce a modified $H$
functional (denoted as $H'$) defined as
%
\begin{equation}\label{eq:CH2}
   H'(t)=\sum_{M=1}^{K}\int f_M(\vec{v},t) \ln f_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%
This $H'$ describes a system with a global energy $E$, a global number of free
classical particles $N$, global volume $V$, and a temperature $T$. Nevertheless,
the global system, in turn, consists of a set of $K$ small cells with a constant
volume $\delta V_M$, whose local homogeneous distribution functions is
$f_{M}(\vec{v},t)$.

@@

To prove the Boltzmann's $H$-theorem (using (\ref{eq:CH2}) as functional), we need
to assume that the distribution function $f_M(\vec{v},t)$ must satisfy the
following assumptions:
%
\begin{itemize}
  \item The local equilibrium hypothesis in each cell ($f_{M}$ must be
    spatial-homogeneous for each $M$).
  \item The non-homogeneous distribution function assumption in the total
    volume ($f_{M}$ must be different among all cells).
  \item The following restrictions
    %
    \begin{equation}\label{restrictionoutsideclassical}
      \int f_{M}(\vec{v},t) \mathrm{d}^{3}v=
      \bar{\mathcal{N}}+\Delta_M(t); \ \ \ \ 
      \int f_{M}(\vec{v},t) \epsilon(\vec{v}) \mathrm{d}^{3}v=
      \bar{\mathcal{E}}+ \delta_M(t),
    \end{equation}
    %
    where $\bar {\mathcal{N}}$ and $\bar{\mathcal{E}}$ are
    the local particle number and the local energy in equilibrium
    %
    \begin{equation}
      \bar{\mathcal{N}}=
      \int \bar{f}(\vec{v}) \mathrm{d}^{3}v ; \quad
      \bar{\mathcal{E}}=
      \int \bar{f}(\vec{v})\epsilon(\vec{v}) \mathrm{d}^{3}v,
    \end{equation}
    %
    and $\Delta_M$, $\delta_M$ could be seen as a deviation from
    $\bar{\mathcal{N}}$ and $\bar{\mathcal{E}}$ respectively,
    with $\Delta_M(t)\ll\bar{\mathcal{N}}$ and
    $\delta_M(t)\ll\bar{\mathcal{E}}$.
\end{itemize} 
%
Those conditions described a system out of equilibrium but no so far from it,
that is a perturbed system and will suffer a \textit{relaxation process, i.e.}
The time evolution of a perturbated system to the equilibrium state.

@@

Joining both results, we can say the following statement:
Consider a classical gas in a total volume $V$ (divided in $K$ cells of equal
volume elements), total energy $E$ and total number of free particles $N$.
Consider also the system has inhomogeneities and suffers a relaxation process.
If we define the following functional
%
\begin{equation}\label{CH3}
   H'(t)=\sum_{M=1}^{K}\int f_M(\vec{v},t) \ln f_M(\vec{v},t)\mathrm{d}^3v,
\end{equation}
%
where $f_M(\vec{v},t)$ is the local distribution function of each cell in the
system, then in the first-order approximation
%
\begin{equation}
    \frac{dH'}{dt} \leq 0.
\end{equation}
%
This statement will be the classical $H$-theorem with inhomogeneities.

In the next section, we present a proposal quantum version of the
$H$-theorem defined by Tolman and our proposal quantum version of the $H$
theorem applying the method of the volume divided into cells.

### 2020-12-16 tex/disscussion.tex
Typos, 1st intro paragraph, secs. 2.0, 2.1

\section{Discussions and remarks}

In the literature, the Boltzmann's $H$ theorem is proved using a spatially homogeneous distribution function. In this work, we used the local equilibrium hypothesis and as a consequence of this, it was necessary to include the homogeneous distribution function hypothesis. However, we included inhomogeneities in the system through the division of the volume $V$ in cells and the non-homogeneous distribution function assumption in the total volume. Moreover, the Boltzmann's $H$ theorem still holds when the system includes inhomogeneities, only if the local equilibrium hypothesis holds. We can include the fact that we proved the Boltzmann's $H$ theorem without using the Boltzmann's transport equation. It was only necessary to specify the behavior of the deviation $g$ and its derivative in time. Also, we obtained the Boltzmann's transport equation in equilibrium from the Boltzmann's $H$ functional through the variational method. Nevertheless, in order to obtain the complete Boltzmann's transport equation, it was necessary to include the collision term because of the variation of the $H$ functional is not zero when the system is out of equilibrium. This is a consequence of the $H$ theorem.\\

@@

In the quantum version, we used the same assumptions to prove the $H$ theorem, those were the behavior of the deviation $g_{nM}$ and its derivative, the local equilibrium hypothesis, and the non-homogeneous distribution function assumption. With the proof of the quantum version of the $H$ theorem, we can say that a quantum gas inside a volume $V$ that is out of equilibrium with inhomogeneities, in the first-order approximation the system evolves to the equilibrium state as long as the local equilibrium hypothesis holds. In addition to this, the system is in the equilibrium state when the derivative of the variational entropy with respect to time is equal to zero, and as a consequence of this, the distribution function corresponds to the Bose-Einstein or Fermi-Dirac distribution.\\

@@

According to the classical analysis of the $H$ theorem, the expected behavior of the distribution function is presented in the system if the molecular chaos hypothesis holds, but in a meantime, called \textit{relaxation time}, the distribution function behaves as we expect, such as the distribution function evolves to the equilibrium state keeping that $\frac{dH}{dt}\leq 0$.\\

>>>>

\section{Discussions and remarks}\label{sec:disscussion}

In the literature, the Boltzmann's $H$-theorem is proved using a spatially homogeneous distribution function. In this work, we used the local equilibrium hypothesis and as a consequence of this, it was necessary to include the homogeneous distribution function hypothesis. However, we included inhomogeneities in the system through the division of the volume $V$ in cells and the non-homogeneous distribution function assumption in the total volume. Moreover, the Boltzmann's $H$-theorem still holds when the system includes inhomogeneities, only if the local equilibrium hypothesis holds. We can include the fact that we proved the Boltzmann's $H$-theorem without using the Boltzmann's transport equation. It was only necessary to specify the behavior of the deviation $g$ and its derivative in time. Also, we obtained the Boltzmann's transport equation in equilibrium from the Boltzmann's $H$ functional through the variational method. Nevertheless, in order to obtain the complete Boltzmann's transport equation, it was necessary to include the collision term because of the variation of the $H$ functional is not zero when the system is out of equilibrium. This is a consequence of the $H$-theorem.\\

@@

In the quantum version, we used the same assumptions to prove the $H$-theorem, those were the behavior of the deviation $g_{nM}$ and its derivative, the local equilibrium hypothesis, and the non-homogeneous distribution function assumption. With the proof of the quantum version of the $H$-theorem, we can say that a quantum gas inside a volume $V$ that is out of equilibrium with inhomogeneities, in the first-order approximation the system evolves to the equilibrium state as long as the local equilibrium hypothesis holds. In addition to this, the system is in the equilibrium state when the derivative of the variational entropy with respect to time is equal to zero, and as a consequence of this, the distribution function corresponds to the Bose-Einstein or Fermi-Dirac distribution.\\

@@

According to the classical analysis of the $H$-theorem, the expected behavior of the distribution function is presented in the system if the molecular chaos hypothesis holds, but in a meantime, called \textit{relaxation time}, the distribution function behaves as we expect, such as the distribution function evolves to the equilibrium state keeping that $\frac{dH}{dt}\leq 0$.\\

### 2020-12-16 tex/introduction.tex
Typos, 1st intro paragraph, secs. 2.0, 2.1

In the Kinetic Theory of dilute gases, the behavior of a dilute classical gas
was described by Boltzmann through the Boltzmann transport equation.
Furthermore, the evolution of an equilibrium state for any system was proved by
means the Boltzmann's $H$ theorem (or classical $H$ theorem) and the
spatial-homogeneous distribution function hypothesis.
However, there are some aspects in the classical $H$ theorem and the
Boltzmann's $H$ functional that deserve some review to accomplish better
consistency, for example, the modification of the $H$ theorem in stochastic
trajectories, violations in the second law of thermodynamics, and the relation
between Shannon's measure of information and the Boltzmann's entropy %[6-10].
\cite{bib:nemilov2018,bib:keizer1987,bib:onsager1931,bib:brown2008,bib:dragoljub2009}.

To obtain the well-known correspondence principle between classical mechanics
and quantum mechanics, a quantum version of the Boltzmann's $H$ theorem (or
quantum $H$ theorem) and the quantum analogous to the Boltzmann transport
equation is necessary such that in the appropriate limit, the  classical $H$
theorem and the Boltzmann transport equation should be recovered.

One of the first developments in this field was developed by Tolman
\cite{bib:tolman}. He proposes one of the first quantum versions of the
Boltzmann's $H$ theorem using the probability transition relation, the random
phases hypothesis, and a homogeneous distribution function. The equation of
motion for the occupation number (necessary to prove the quantum $H$ theorem in
the same way that Boltzmann's procedure) was obtained by applying the time
perturbation theory. However, to obtain the flux relations in analogy to the
Boltzmann's method, a quantum version of the transport equation that includes a
non-homogeneous distribution function is required.

On the other hand, some works in the quantum operator formalism attempt to
describe quantum transport phenomena using the Hamiltonian of the system and
the master equation (analogous to the Boltzmann transport equation)
\cite{bib:gorban2014,bib:bennaim2017,bib:tolman,bib:li2019}. %[12-15]
Besides, some authors construct the $H$ functional and develop the proof of the
quantum $H$ theorem \cite{bib:silva2010,bib:deroeck2006,bib:grabert1974}.
However, the homogeneous distribution function hypothesis and the
correspondence principle are uncleared or not discussed. Moreover, the validity
of the quantum $H$ theorem and the second law of thermodynamics, and the
interpretation of the quantum entropy is still in discussion %[19-26].
\cite{bib:silva2010,bib:deroeck2006,bib:grabert1974,bib:acharya2019,%
bib:kastner2017,bib:gring2012,bib:han2015,bib:das2018}.

In order to contribute to the construction of a consistent classical and
quantum $H$ theorem, and a new formalism to describe out of equilibrium
systems, we develop a new theoretical basis to describe them. In particular, we
are interested in including a non-homogeneous distribution function in the
mathematical model, and so including non-homogeneous systems in the $H$
theorem. Besides, we are interested in developing the quantum analogous of
Boltzmann's $H$ theorem considering a non-spatial-homogeneous distribution
function and solve the problem of the correspondence principle mentioned above.

The first section of this work consists of a review of the
Boltzmann transport equation, the Boltzmann $H$ theorem, the Maxwell-Boltzmann
distribution, and our alternative form to prove the classical $H$ theorem. The
second part consists of a review of the proposal from Tolman about the quantum
version of the $H$ theorem, the Bose-Einstein and Fermi-Dirac distributions,
and our alternative form to prove the quantum version of the $H$ theorem. The
third section consists of a classical-quantum correspondence between the
quantum $H$ functional and the classical $H$ functional. Finally, the last
section consists of a proposal to explore relaxation processes in the quantum
case obtaining the analogous quantum Boltzmann transport equation.

>>>>

In the Kinetic Theory of dilute gases, the behavior of a dilute classical gas
was described by Boltzmann through the Boltzmann transport equation.
Furthermore, the evolution of an equilibrium state for any system was proved by
means the Boltzmann's $H$-theorem (or classical $H$-theorem) and the
spatial-homogeneous distribution function hypothesis.
However, there are some aspects in the classical $H$-theorem and the
Boltzmann's $H$ functional that deserve some review to accomplish better
consistency, for example, the modification of the $H$-theorem in stochastic
trajectories, violations in the second law of thermodynamics, and the relation
between Shannon's measure of information and the Boltzmann's entropy %[6-10].
\cite{bib:nemilov2018,bib:keizer1987,bib:onsager1931,bib:brown2008,bib:dragoljub2009}.

To obtain the well-known correspondence principle between classical mechanics
and quantum mechanics, a quantum version of the Boltzmann's $H$-theorem (or
quantum $H$-theorem) and the quantum analogous to the Boltzmann transport
equation is necessary such that in the appropriate limit, the  classical $H$
theorem and the Boltzmann transport equation should be recovered.

One of the first developments in this field was developed by Tolman
\cite{bib:tolman}. He proposes one of the first quantum versions of the
Boltzmann's $H$-theorem using the probability transition relation, the random
phases hypothesis, and a homogeneous distribution function. The equation of
motion for the occupation number (necessary to prove the quantum $H$-theorem in
the same way that Boltzmann's procedure) was obtained by applying the time
perturbation theory. However, to obtain the flux relations in analogy to the
Boltzmann's method, a quantum version of the transport equation that includes a
non-homogeneous distribution function is required.

On the other hand, some works in the quantum operator formalism attempt to
describe quantum transport phenomena using the Hamiltonian of the system and
the master equation (analogous to the Boltzmann transport equation)
\cite{bib:gorban2014,bib:bennaim2017,bib:tolman,bib:li2019}. %[12-15]
Besides, some authors construct the $H$ functional and develop the proof of the
quantum $H$-theorem \cite{bib:silva2010,bib:deroeck2006,bib:grabert1974}.
However, the homogeneous distribution function hypothesis and the
correspondence principle are uncleared or not discussed. Moreover, the validity
of the quantum $H$-theorem and the second law of thermodynamics, and the
interpretation of the quantum entropy is still in discussion %[19-26].
\cite{bib:silva2010,bib:deroeck2006,bib:grabert1974,bib:acharya2019,%
bib:kastner2017,bib:gring2012,bib:han2015,bib:das2018}.

In order to contribute to the construction of a consistent classical and
quantum $H$-theorem, and a new formalism to describe out of equilibrium
systems, we develop a new theoretical basis to describe them. In particular, we
are interested in including a non-homogeneous distribution function in the
mathematical model, and so including non-homogeneous systems in the $H$
theorem. Besides, we are interested in developing the quantum analogous of
Boltzmann's $H$-theorem considering a non-spatial-homogeneous distribution
function and solve the problem of the correspondence principle mentioned above.

The organization of this article is as follows.
In Section~\ref{sec:classical}, we highlight, to our purposes, the most
fundamental assumptions required to proof the Boltzmann $H$-theorem,
we provide an alternative method to obtain the Maxwell-Boltzmann
distribution using the variational method, and an alternative demonstration of
the $H$-theorem for classical systems. In Section~\ref{sec:quantum}
we review the Tolman proposal for the quantum
version of the $H$-theorem (quantum $H$-theorem)
\commr{and how the Bose-Einstein and Fermi-Dirac distributions are treated within
this framework},
and our alternative proof of the quantum $H$-theorem. In 
Section~\ref{sec:qccorrespondence}
the classical-quantum correspondence between the
quantum $H$ functional and the classical $H$ functional is analyzed.
In Section~\ref{sec:relaxproc} we explore how relaxation processes occur
in a quantum ideal gas; here we provide a time-evolution equation that could be
considered the analogous of the classical Boltzmann transport equation. Finally,
we discuss some key ideas derived from our approach in Section~\ref{sec:disscussion},
and provide some conclusions in Section~\ref{sec:conclusions}.

### 2020-12-16 tex/quantum.tex
ssh test
Typos, 1st intro paragraph, secs. 2.0, 2.1

\section{Quantum scheme}

@@

The $H$ functional in quantum mechanics presented by Tolman \cite{bib:tolman}
is defined as
%
\begin{equation}
    H=\sum_i n_i \ln n_i -(n_i\pm g_i)\ln (g_i \pm n_i)\pm g_i\ln g_i, \label{quantumh}
\end{equation}
%
where $n_i$ is the occupation number and $g_i$ is an energy cell. Also, an
equation for the rate of change in the number of particles in group $\kappa$ is
%
\begin{eqnarray}
    \frac{d n_{\kappa}}{dt}&=&-\sum_{\lambda,(\mu \nu)}A_{\kappa\lambda,\mu\nu} n_{\kappa}n_{\lambda}(g_{\mu}\pm n_{\mu})(g_{\nu}\pm n_{\nu})\nonumber \\
    &&+\sum_{\lambda,(\mu \nu)}A_{\mu\nu,\kappa\lambda} n_{\mu}n_{\nu}(g_{\kappa}\pm n_{\kappa})(g_{\lambda}\pm n_{\lambda}),\label{changen}
\end{eqnarray}
%
where $A_{\kappa\lambda,\mu\nu}$ is a tensor expressed as
%
\begin{equation}
  A_{\kappa\lambda,\mu\nu}=\frac{4\pi^{2}}{h}\frac{|I_1\pm I_2|^2}{\Delta \epsilon},
\end{equation}
%
where $\epsilon$ is the energy and $|I_1-I_2|^2=|V_{mn,kl}|^2$ where
$V_{mn,kl}$ is the perturbation matrix of binary particles. Those expressions
are obtained by the perturbation theory.

@@

%\subsection{The quantum analogous of the Boltzmann's $H$ theorem}
We rewrite the \textit{variational entropy} (\ref{entropy}) as
%
\begin{equation}
    \Ss(t)=\sum_{M=1}^{K} \sum_{n} \left[ f_{Mn} \ln f_{Mn} \pm (1 \mp f_{Mn}) \ln (1 \mp f_{Mn}) \right] \de V_M \delta \epsilon_n \label{entropy2},
\end{equation}
%
where, for simplicity, we have defined $f(\epsilon_{n},t)\equiv f_{Mn}$.

@@

When the system is in equilibrium, $g_{nM}=0$ and in consequence of that, $\dot
g_{nM}=0$ and finally from (\ref{cambioH1}) $\frac{d\Ss}{dt}=0$. Then, we can
say that any ideal quantum gas perturbed always evolves to the equilibrium
state, such as the classical $H$ theorem. Then, we obtained the quantum version
of the Boltzmann's $H$ theorem including inhomogeneities in the system.

>>>>

\section{Quantum scheme}\label{sec:quantum}

@@

The $H$ functional in quantum mechanics presented by Tolman \cite{bib:tolman}
is defined as
%
\begin{equation}\label{quantumh}
    H=\sum_i n_i \ln n_i -(n_i\pm g_i)\ln (g_i \pm n_i)\pm g_i\ln g_i, 
\end{equation}
%
where $n_i$ is the occupation number and $g_i$ is an energy cell. Also, an
equation for the rate of change in the number of particles in group $\kappa$ is
%
\begin{eqnarray}
    \frac{d n_{\kappa}}{dt}&=&-\sum_{\lambda,(\mu \nu)}A_{\kappa\lambda,\mu\nu} n_{\kappa}n_{\lambda}(g_{\mu}\pm n_{\mu})(g_{\nu}\pm n_{\nu})\nonumber \\
    &&+\sum_{\lambda,(\mu \nu)}A_{\mu\nu,\kappa\lambda} n_{\mu}n_{\nu}(g_{\kappa}\pm n_{\kappa})(g_{\lambda}\pm n_{\lambda}),\label{changen}
\end{eqnarray}
%
where $A_{\kappa\lambda,\mu\nu}$ is a tensor expressed as
%
\begin{equation}
  A_{\kappa\lambda,\mu\nu}=\frac{4\pi^{2}}{h}\frac{|I_1\pm I_2|^2}{\Delta \epsilon},
\end{equation}
%
where $\epsilon$ is the energy and $|I_1-I_2|^2=|V_{mn,kl}|^2$ where
$V_{mn,kl}$ is the perturbation matrix of binary particles. Those expressions
are obtained by the perturbation theory.

@@

%\subsection{The quantum analogous of the Boltzmann's $H$-theorem}
We rewrite the \textit{variational entropy} (\ref{entropy}) as
%
\begin{equation}
    \Ss(t)=\sum_{M=1}^{K} \sum_{n} \left[ f_{Mn} \ln f_{Mn} \pm (1 \mp f_{Mn}) \ln (1 \mp f_{Mn}) \right] \de V_M \delta \epsilon_n \label{entropy2},
\end{equation}
%
where, for simplicity, we have defined $f(\epsilon_{n},t)\equiv f_{Mn}$.

@@

When the system is in equilibrium, $g_{nM}=0$ and in consequence of that, $\dot
g_{nM}=0$ and finally from (\ref{cambioH1}) $\frac{d\Ss}{dt}=0$. Then, we can
say that any ideal quantum gas perturbed always evolves to the equilibrium
state, such as the classical $H$-theorem. Then, we obtained the quantum version
of the Boltzmann's $H$-theorem including inhomogeneities in the system.
